\bf Model Visualization Techniques for a Social Network Model

### 2018-05-29 paper-child.Rnw
still needs some trimming :scissors:

A key component of each step of the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$
This simulation process follows the steps of the continuous-time Markov chain. \st{Each tie change in the CTMC is referred to as one ``microstep". At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations. We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

<<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes. The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>% group_by(period, rep) %>%
  mutate(ms = row_number()) %>%
  filter(ms == 1, period == 1) %>%
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>%
  ungroup() %>%
  group_by(ego, alter) %>%
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>%
  group_by(ego) %>%
  mutate(csum = cumsum(alterProb)) %>%
  arrange(ego, csum) %>%
  mutate(totego = sum(count),
         empprob = count / totego) -> cumprobs
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process. We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model. %By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$. Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network. For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states. It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases. 
Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes. The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made. The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change. In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected. There are, however, many steps that are never taken. Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties. On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.  It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more. In addition, we can see that the ties changed most often are removing ties, not adding ties. This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.

@@

Another way to view these transition probabilities is through the adjacency matrix visualization. In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse: %showing the lack of coverage in the model.
of the $16^2 = 256$ possible steps for the CTMC to take, only 103, or about 40\%, are taken in the 1,000 simulated chains. \st{This reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken. This effect could only be exacerbated as more steps are taken in the CTMC, leading to a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.}

<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5>>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
ansnullchains %>%
  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  #dplyr::group_by(rep) %>%
  dplyr::select(rep, from = ego, to = alter) %>%
  dplyr::mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to),
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>%
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>%
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave. The number of steps taken from wave 1 to wave 2 varies. In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103. In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8. In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right. In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way. We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again. There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation. \st{So, even though the CTMC makes about the right \textit{number} of changes as it was designed to do, the changes it is making are not necessarily in the right direction.}

@@

We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}. We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations. Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later. We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000. Finally, we visualize all this information together in Figure ~\ref{fig:alledgests}. In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences of the edges which appear in the second wave of data. But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network. \st{This solidifies what we found in Figures~\ref{fig:histogramalters} and~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.} %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization. The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges. This makes sense, because we condition on the first wave, but it would be better if the model explored the space of possible edges more than it currently does. It could be that the simple model M1 is not complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs ``get stuck" when exploring the parameter space, but we cannot know this .}

@@

\hh{The visualizations should address most elements of structure described before, i.e. follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }

@@

\st{XX HH help please!}

>>>>

A key component of each step of the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$
This simulation process follows the steps of the continuous-time Markov chain. Each tie change in the CTMC is referred to as one ``microstep". At each microstep, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations. We want to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

<<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes. The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.", eval = FALSE>>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>% group_by(period, rep) %>%
  mutate(ms = row_number()) %>%
  filter(ms == 1, period == 1) %>%
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>%
  ungroup() %>%
  group_by(ego, alter) %>%
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>%
  group_by(ego) %>%
  mutate(csum = cumsum(alterProb)) %>%
  arrange(ego, csum) %>%
  mutate(totego = sum(count),
         empprob = count / totego) -> cumprobs
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process. We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model. %By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$. Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network. For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states. It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases. 
Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
%In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes. The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made. The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change. In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected. There are, however, many steps that are never taken. Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties. On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.  It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more. In addition, we can see that the ties changed most often are removing ties, not adding ties. This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.

@@

%Another way to view these transition probabilities is through the adjacency matrix visualization. 
In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse: %showing the lack of coverage in the model.
of the $16^2 = 256$ possible steps for the CTMC to take, only 103, or about 40\%, are taken in the 1,000 simulated chains. This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken. This effect could only be exacerbated as more steps are taken in the CTMC, leading to 
a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.

<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5, eval = FALSE>>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
ansnullchains %>%
  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  #dplyr::group_by(rep) %>%
  dplyr::select(rep, from = ego, to = alter) %>%
  dplyr::mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to),
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>%
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>%
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

%We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave. The number of steps taken from wave 1 to wave 2 varies. In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103. In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8. In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right. In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way. We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again. There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation. \st{So, even though the CTMC makes about the right \textit{number} of changes as it was designed to do, the changes it is making are not necessarily in the right direction.}

@@

Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}. To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations. Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later. We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000. Finally, we visualize all this information together in Figure~\ref{fig:alledgests}. In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences of the edges which appear in the second wave of data. But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network. This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism. %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization. The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges. This makes sense, because we condition on the first wave, but it would be better if the model explored the space of possible edges more than it currently does. It could be that the simple model M1 is not complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs ``get stuck" when exploring the parameter space, but we cannot know this .}

@@

%\hh{The visualizations should address most elements of structure described before, i.e. follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }

@@
